An examination of the high-order dynamic interactions underlying multi-dimensional timeseries data

### 2018-10-16 paper/main.tex
checking in changes-- levelup [WIP]

\subsection*{Dynamic correlations}
Given a matrix of observations, we can compute the (static)
correlations between any pair of observations, $\mathbf{X}_i$ and
$\mathbf{X}_j$ using:
\begin{align}
  \mathrm{corr}(\mathbf{X}_i, \mathbf{X}_j) &= \frac{\sum_{t=1}^T \left(\mathbf{X}_i(t)
                                              -
                                              \hat{\mathbf{X}_i}\right) \left(\mathbf{X}_j(t)
                                              -
                                              \hat{\mathbf{X}_j}\right)}{\sqrt{\sum_{t=1}^T
                                              \sigma^2_{\mathbf{X}_i} 
                                              \sigma^2_{\mathbf{X}_j}}},~\mathrm{where}\\\label{eqn:corr}
  \hat{\mathbf{X}_k} &= \sum_{t=1}^T
                       \mathbf{X}_k(t),~\mathrm{and}\\
  \sigma^2_{\mathbf{X}_k} &= \sum_{t=1}^T \left( \mathbf{X}_k -
                            \hat{\mathbf{X}_k} \right)^2 
\end{align}

@@

\subsection*{Higher-order correlations}
Given a timeseries of dynamic correlations (e.g., obtained using
Eqn.~\ref{eqn:timecorr}), higher-order correlations reflect the
dynamic correlations between columns of $\mathbf{Y}$.  Given unlimited
computing resources, one could use repeated applications of
Equation~\ref{eqn:timecorr} to estimate these higher-order
correlations (i.e., substituting in the previous output, $\mathbf{Y}$,
for the input, $\mathbf{X}$ in the equation).  However, because each output
$\mathbf{Y}$ has $O(F^2)$ columns relative to $F$ columns in the input
$\mathbf{X}$, the output of Equation~\ref{eqn:timecorr} grows with the
square of the number of repeated applications.  When $F$ is large,
this approach quickly becomes intractable.

@@

\subsection*{Timepoint decoding}

\subsection*{Prediction}

\section*{Results}
%currently Lucy, Kirsten, and Paxton have agreed to make figures

>>>>

\subsection*{Dynamic correlations}
Given a matrix of observations, we can compute the (static)
correlations between any pair of observations, $\mathbf{X}_i$ and
$\mathbf{X}_j$ using:
\begin{align}
  \mathrm{corr}(\mathbf{X}_i, \mathbf{X}_j) &= \frac{\sum_{t=1}^T \left(\mathbf{X}_i(t)
                                              -
                                              \bar{\mathbf{X}_i}\right) \left(\mathbf{X}_j(t)
                                              -
                                              \bar{\mathbf{X}_j}\right)}{\sqrt{\sum_{t=1}^T
                                              \sigma^2_{\mathbf{X}_i} 
                                              \sigma^2_{\mathbf{X}_j}}},~\mathrm{where}\\\label{eqn:corr}
  \bar{\mathbf{X}_k} &= \sum_{t=1}^T
                       \mathbf{X}_k(t),~\mathrm{and}\\
  \sigma^2_{\mathbf{X}_k} &= \sum_{t=1}^T \left( \mathbf{X}_k -
                            \bar{\mathbf{X}_k} \right)^2 
\end{align}

@@

\subsection*{Higher-order correlations}
Given a timeseries of dynamic correlations (e.g., obtained using
Eqn.~\ref{eqn:timecorr}), higher-order correlations reflect the
dynamic correlations between columns of $\mathbf{Y}$.  Given unlimited
computing resources, one could use repeated applications of
Equation~\ref{eqn:timecorr} to estimate these higher-order
correlations (i.e., substituting in the previous output, $\mathbf{Y}$,
for the input, $\mathbf{X}$ in the equation).  However, because each
output $\mathbf{Y}$ has $O(F^2)$ columns relative to $F$ columns in
the input $\mathbf{X}$, the output of Equation~\ref{eqn:timecorr}
grows with the square of the number of repeated applications (total
cost of computing $n$\textsuperscript{th} order correlations is
$O(F^{2(n - 1)})$ for $n \in \mathcal{J}, n > 0$).  When $F$ or $n$ is large,
this approach quickly becomes intractable.

To make progress in computing $\mathbf{Y_{n+1}}$, we can approximate
$\mathbf{Y}_n$ by computing an $O(F)$-dimensional embedding of
$\mathbf{Y}_n$, termed $\hat{\mathbf{Y}}_n$, and then we can apply
Equation~\ref{eqn:timecorr} to $\hat{\mathbf{Y}}_n$ rather than
directly to $\mathbf{Y}_n$.  This enables us to maintain $O(n)$
scaling with respect to $n$, rather than exponential scaling via the
direct approach.

There are many possible methods for computing $\hat{\mathbf{Y}}_n$
from $\mathbf{Y}_n$, including traditional dimensionality reduction
approaches and graph theory based approaches as described next.  In the
\textit{Discussion} section we elaborate on other potential approaches.

\subsubsection*{Dimensionality reduction-based approaches to computing
  $\hat{\mathbf{Y}}_n$}

Commonly used dimensionality reduction algorithms include Principal
Components Analysis~\citep[PCA; ][]{Pear01}, Probabilistic
PCA~\citep[PPCA; ][]{TippBish99}, Exploratory Factor
Analysis~\citep[EFA; ][]{Spea04}, Independent Components
Analysis~\citep[ICA; ][]{JuttHera91, ComoEtal91}, $t$-Stochastic
Neighbor Embedding~\citep[$t$-SNE; ][]{MaatHint08}, Uniform Manifold
Approximation and Projection~\citep[UMAP; ][]{McInHeal18},
non-negative matrix factorization~\citep[NMF; ][]{LeeSeun99},
Topographic Factor Analysis (TFA)~\cite{MannEtal14b}, Hierarchical
Topographic Factor analysis (HTFA)~\cite{MannEtal18}, Topographic
Latent Source Analysis (TLSA)~\cite{GersEtal11}, Dictionary
learning~\citep{MairEtal09a, MairEtal09b}, deep
autoencoders~\citep{HintSala06}, among others.  While complete
characterizations of each of these algorithms is beyond the scope of
the present manuscript, the general intuition driving these
approaches is to compute the $\hat{\mathbf{Y}}$ with $i$ columns that
is closest to the original $\mathbf{Y}$ with $j$ columns, and where
(typically) $i \ll j$.  The different approaches place different
constraints on what properties $\hat{\mathbf{Y}}$ must satisfy and
which aspects of the data are compared (and how) to characterize the
match between $\hat{\mathbf{Y}}$ and  $\mathbf{Y}$.

\subsubsection*{Graph theory-based approaches to computing
  $\hat{\mathbf{Y}}_n$}

\textbf{JRM TODO: Need references for each of the below, plus citation
  of \href{https://github.com/aestrivex/bctpy/wiki}{bctpy toolbox} (source of measures listed below)}
\begin{itemize}
\item Betweenness centrality: Wei; Node Betweenness; edge betweenness
  centrality; edge betweeness wei; 
\item Shannon-entropy based diversity coefficient
\item Eigenvector centrality, pagerank centrality
\item flow coefficient (Honey et al. 2007, PNAS)
\item k-coreness centrality (directed, undirected)
\item within-module degree
\item participation coefficient (signed, unsigned)
\item subgraph centrality
\end{itemize}

Idea: for each timepoint, compute relevant measure for node.  Result:
a $T$ by $F$ matrix characterizing how the given measure for each node
over time.  We can use this as $\hat{\mathbf{Y}}_n$ and then apply
Equation~\ref{eqn:timecorr} to compute $\hat{\mathbf{Y}}_{n+1}.$

@@

\subsection*{Timepoint decoding}
Cite HTFA paper, summarize relevant methods.

\subsection*{Prediction}
Use Eqn. 12 from SuperEEG paper, extend notation to treat
$\hat{\mathbf{Y}}_{1...n}(1...(t-1))$ as ``observed'' and use this to
predict $\mathbf{X}(t)$.  Key question: what is the predictive power
added by each new level.

\section*{Results}
\subsubsection*{Synthetic data}
\subsubsection*{Neuroimaging dataset~\citep{SimoEtal16}}
